\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-02-24 src/020chap2.tex
All content done except acknowledgement. TBD: - spell check - unusual whitespaces, wandering floats

The chapter is organized as follows. Section~\ref{sec:copgraphtheory}
discusses the close connections of matrices with \COP in graph
theory. Section~\ref{sec:surveycoptest} discusses the important
results in the area of \cop testing (COT) -- we describe the history
of this problem's several solutions including Tucker's forbidden
submatrices, \PQtree, \PQRtree and ICPIA. \tnote[dire]{mention the section
  that compares pqr and gen pq tree} 
Section~\ref{sec:surveycopopt} briefly touches upon the scenario of
algorithmically verifying that a matrix indeed does not have \COP. Its
significance lies in the fact that almost all \COT algorithms, upon
detection of no \COP, simply reports the same without any certificate
for verification during algorithm implementation. The two approaches
presented in this section gives a certificate.

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote[pressing]{cite BHY00}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote[pressing]{cite Ben59,
      Haj57}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65} (This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.)
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.
    \end{enumerate}
    The set of unit interval graphs and the set of proper interval
    graphs coincide \cite{rob69, gar07}.
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using
the result \cite{hmpv00} \cite[Th~2.7]{d08phd} which demonstrates how
such a derivation can be done. However, this does not necessarily
yield an efficient algorithm. We will see results that directly solve
the problem on matrices since it is known that
questions~(\ref{q:testcop}) and (\ref{q:copperm}) stated above for
\COP are efficiently solvable.  Table~\ref{tab:cophistory} gives a
snapshot of these results.

@@

A few years later, a deeply significant result based on very different
ideas in understanding \COP came from Tucker which gave a
combinatorial (negative) characterization of matrices with COP
\cite{at72}. This result influenced most of the \COP results that
followed in the literature including linear time
algorithms\tnote[pressing]{linear? did it? which ones?} for \COP
recognition.

@@

\subsection{\PQRtree\ --  COP for set systems}
\label{sec:surveycertalgo}

@@

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the {\em \cop}
  if there exists a linear order or permutation $\sigma = w_1w_2\ldots
  w_n$ that can be applied to $U$ \stt each set $S \in \cF$ becomes a
  consecutive subsequence (slso termed an {\em interval})
  $w_{i}w_{i+1}\ldots w_{i+k-1}$ on $\sigma$ for some positive integer
  $i \le n+1-k$ where $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

  \item {\em Orthogonal sets of a set system} $\cF$ with universe $U$
    are subsets of $U$ that are orthogonal to all sets in $\cF$. The
    set of all orthogonal sets to $\cF$ is denoted by
    $\cF^{\bot}$\footnote{\cite[Def.~3.1]{mcc04} uses the term {\em non-overlapping
    family} of $\cF$ and denotes it by $\cN(\cF)$. \label{mcc1}}.

  \item $\cF$ is called {\em complete}\footnote{\cite{mcc04} calls
      this a {\em weakly partative family}. \label{mcc2}} if the following hold
    true for every pair of non-orthogonal\footnote{Or overlapping.} sets
    $A, B$ in $\cF$.
    \begin{enumerate}[i.]
    \item $\cT(U) \subset \cF$
    \item $A \cup B \in \cF$
    \item $A \cap B \in \cF$
    \item $A \setminus B \in \cF$
    \item $B \setminus A \in \cF$
    \end{enumerate}
    In other words, $\cF$ contains all the trivial
    subsets of $U$, $A \cup B$ and the partitions of $A \cup B$
    defined by intersection and set difference.

@@

Generalized \PQtree or \gPQtree is a data structure defined in
\cite{n89} to represent all orthogonal sets of a set system $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system $\cF$
\tnote[pressing]{define substitutio decomp.} and subsequently \cite{mcc04} gives
a new characterization of $\cF$ using a so-called incompatibility
graph. \PQRtree is defined in \cite{mm96} as a data
structure to represent any set system $\cF$ with additional
information in their $R$-nodes if $\cF$ has no \COP.  All three of
these data structures are proposed as generalizations of \cite{bl76}'s
\PQtree and hence produce the \PQtree if $\cF$ has \COP. As a whole,
all these three data strutures are largely identical with their
differences being notional.  In this section, we will discuss the
basic theory that they all hold. We will predominantly use the
terminology from \cite{mm96} and refer to the data structure as
\PQRtree.

@@

\tnote[dire]{narrative is abrupt from here. add some explanation for the
  theorem and proposition below?}

@@

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree, as alluded to by
Corollary~\ref{cor:pqrrecursive}. They are calculated as follows.
First method is by computing the overlap components of $\cF$. The
overlap components is the partition that results from the {\em
  overlap} equivalence relation which is nothing but {\em
  non-orthogonal} equivalence relation (It is easy to verify
  that this relation is indeed an equivalence relation.). Overlap
components are linearly computable\cite{mm95,wlh92}. Once these
elements are factored out, the rest of the node sets are obtained by
the second method which is identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01,mm96-ref11} and even in
logspace\cite{kklv10}.

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote[pressing]{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the above the three cases as {\em
  prime}, {\em linear} and {\em degenerate}
respectively. \cite{mcc04}'s generalized \PQtree is created in similar
ways as \PQRtree above. This tree is, in essence, the Hasse diagram of
what they call {\em strong elements} of weak closure,
$\cW(\cF)$(\ie $\overline{\cF}$). Strong elements of a set
family are elements that do not overlap with any other elements in the
family, \ie it is orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

As mentioned before, we observe the three data strutures of \PQRtree,
\gPQtree and generalized \PQtree to be equivalent. The terminologies
of the theory of all these data structures is summarised in
Table~\ref{tab:pqrcomparison}.

@@

    &
%     \parbox[t]{\colwidth}
%       {}
    \tnote[dire]{PQR tree def. }
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
      \tnote[dire]{PQR tree def. }
    }
    \\\hline

@@

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }&
    \tnote[dire]{give def of P Q R nodes}
    \\\lasthline

@@

\subsection{ICPIA - a set cardinality based \COP test}
\label{sec:icpiasurvey}
\tnote[dire]{narrative is abrupt}

\cite{nsnrs09} describes a characterization of consecutive-ones
property solely based on the cardinality properties of the sets in the
set system and does not use any variants of \PQtrees.

@@

\subsection{Other \COP testing algorithms}
There are more \COP test algorithms in the literature. We make a
mention of one of them here.  A simple linear time algorithm was
presented by \cite{hmpv00}. They use something called {\em Lex-BFS
  ordering} of the vertices of a graph to decide in linear time
whether the graph is an interval graph. They also show how any
interval graph recognition algorithm can be used to recognize matrices
with \COP. This is shown in Theorem~\ref{th:intgraphcop}.

\tnote[dire]{ abrupt: Add a section closing para. does this method (above)
  lead to efficient algo? }

@@

\subsection{Finding forbidden submatrices}
It is known from Theorem~\ref{th:tuckercop}
that when a matrix does not have \COP, it must have one of Tucker's
forbidden submatrices. However, none of the \COP test algorithms discussed in
Section~\ref{sec:surveycoptest} outputs the forbidden submatrix when
the test is negative. This certainly gives a hint to how hard it is to find
these substructures in the input. \tnote[dire]{Fill neg results on
  finding forbidden submartrices -- boo75, some highlighted stuff from v85}

In this section we will describe \cite{d08phd}'s approximation
algorithm for finding forbidden submatrix. We will use $\Tuckerset$ to
refer to the set of Tucker's forbidden submatrices shown in
Figure~\ref{fig:forbiddensubmatrices}.

@@

\temptext{
\cite[Ch.~3]{d08phd} describes two main algorithms - one is an
approximation algorithm and the other one is an exact algorithm.

\begin{theoremsansproof}[{\cite[Th.~3.1]{d08phd}}] 
\label{th:forbiddenalg}
Let $M$ be a $(*,\Delta)$-matrix of size $m
\times n$. A forbidden submatrix from $\Tuckerset$ in $M$ that has a minimum
number of rows can be found in $O(\Delta^3m^2n\dot(m+n^2))$
time. Within the same time, one can also find a forbidden submatrix
from $\Tuckerset$ in $M$ that has a minimum number of columns, a minimum number
of rows and columns, or a minimum number of entries.

\end{theoremsansproof}

}

\subsection{Incompatibility graph - a certificate for no \COP}

@@

\cite{mcc04} describes another characterization for matrices without
\COP which is efficiently computable. This is done by way of an {\em
  incompatibility graph} of a set system\footnote{Since \COP in
  matrices and set systems are equivalent problems as seen in
  Section~\ref{sec:surveycertalgo} we will use the term set system in
  this section, in line with \cite{mcc04}'s convention.}. The
construction of this graph is based on the following observation.

@@

\begin{definition}[{\em Incompatibility graph \cite[Def.~6.1]{mcc04}}]
  Let $\cF$ be a set system with universe $U$. Let $A_{\cF} = \{(a,b)
  \mid a, b \in U, a \ne b\}$. The {\em incompatibility graph}
  $G_I^\cF$ of $\cF$ is an undirected graph defined as follows.
%  \singlespacing
  \begin{enumerate}
  \item   The vertex set of $G_I^\cF$ is $ A_{\cF}$
  \item  The edge set of $G_I^\cF$ are pairs of the following forms
    \begin{itemize}
    \item $\{(a,b),(b,a)\}$ for all $a, b \in U$
    \item $\{(a,b),(b,c)\}$ if there exists $S \in \cF$ \stt $a, c
      \in S$ and $b \notin S$, for all $a, b, c \in U$
    \end{itemize}
  \end{enumerate}
  \dstop
\end{definition}

Thus an edge between two vertices $(a,b), (b,c)$ in $G_I^{\cF}$ means
that there exists no linear order on $U$ that can place $b$ after $a$
and place $c$ after $b$.  For any \COP order, there must not be any
incompatible pairs. Thus it must consist of an independent set $I$ of
the incompatibility graph. Moreover the independent set must have
exactly half the vertex set of $G_I^\cF$ since the \COP order must
involve all elements in $U$. Secondly, the reverse of a \COP order is
also a \COP order. Thus $A_\cF \setminus I$ must also be an
independent set. In other words, $G_I^\cF$ must be bipartite with
partitions being of equal size if $\cF$ has \COP. An odd cycle is a
forbidden structure in a bipartite graph, hence the same is forbidden
in the incompatibility graph of a set system with \COP. In other
words, if the set system has no \COP, its incompatibility graph must
have an odd cycle.  There is also a requirement on the length of this
odd cycle which is formally stated in the following
theorem. \tnote[pressing]{all refrences to ``following''/abov theorem must be
  chagned to slashrefs. }

>>>>

The chapter is organized as follows. Section~\ref{sec:copgraphtheory}
discusses the close connections of matrices with \COP in graph
theory. Section~\ref{sec:surveycoptest} discusses the important
results in the area of \cop testing (COT) -- we describe the history
of this problem's several solutions including Tucker's forbidden
submatrices, \PQtree, \PQRtree and ICPIA.

It is worth noting that we make a very interesting observation about
\PQRtree, generalised \PQtree and \gPQtree\ -- that the theory that
all these independent data structures are built upon are very similar
and equivalent. This is exhaustively described in
Section~\ref{sec:surveycertalgo} and curiously, this comparison has
not been encountered by us in the literature.

Section~\ref{sec:surveycopopt} briefly touches upon the scenario of
algorithmically verifying that a matrix indeed does not have \COP. Its
significance lies in the fact that almost all \COT algorithms, upon
detection of no \COP, simply reports the same without any certificate
for verification during algorithm implementation. The two approaches
presented in this section gives a certificate.

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \cite{bhy00}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \cite{sb59}.  $G$ is an interval graph \iff its maximal clique matrix
    has COP \cite{fg65} (This follows \cite{gh64} which states
      that the maximal cliques of interval graph $G$ can be linearly
      ordered \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.)
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.
    \end{enumerate}
    The set of unit interval graphs and the set of proper interval
    graphs coincide \cite{rob69, gar07}.
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using
the result \cite{hmpv00} \cite[Th~2.7]{d08phd} which demonstrates how
such a derivation can be done.  They use something called {\em Lex-BFS
  ordering} of the vertices of a graph to decide in linear time
whether the graph is an interval graph. They also show how any
interval graph recognition algorithm can be used to recognize matrices
with \COP. This is shown in Theorem~\ref{th:intgraphcop}.

However, this does not necessarily yield an efficient algorithm. We
will see results that directly solve the problem on matrices since it
is known that questions~(\ref{q:testcop}) and (\ref{q:copperm}) stated
above for \COP are efficiently solvable.  Table~\ref{tab:cophistory}
gives a snapshot of these results.

@@

A few years later, a deeply significant result based on very different
ideas in understanding \COP came from Tucker which gave a
combinatorial (negative) characterization of matrices with COP
\cite{at72}. This result influenced most of the \COP results that
followed in the literature including linear time algorithms for \COP
recognition.

@@

\subsection{\PQRtree\ --  COT for set systems}
\label{sec:surveycertalgo}

@@

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the {\em \cop}
  if there exists a linear order or permutation $\sigma = w_1w_2\ldots
  w_n$ that can be applied to $U$ \stt each set $S \in \cF$ becomes a
  consecutive subsequence (also termed an {\em interval})
  $w_{i}w_{i+1}\ldots w_{i+k-1}$ on $\sigma$ for some positive integer
  $i \le n+1-k$ where $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

  \item {\em Orthogonal sets of a set
      system}\footnote{\cite[Def.~3.1]{mcc04} uses the term {\em
        non-overlapping family} of $\cF$ and denotes it by
      $\cN(\cF)$. \label{mcc1}} $\cF$ with universe $U$ are subsets of
    $U$ that are orthogonal to all sets in $\cF$. The set of all
    orthogonal sets to $\cF$ is denoted by $\cF^{\bot}$.

  \item $\cF$ is called {\em complete}\footnote{\cite{mcc04} calls
      this a {\em weakly partative family}. \label{mcc2}} if the
    following hold true for every pair of non-orthogonal (overlapping)
    sets $A, B$ in $\cF$.

    \begin{tabular}[h]{lllll}
%    \begin{enumerate}[i.]
    (i) $\cT(U) \subset \cF$ &
    (ii) $A \cup B \in \cF$ &
    (iii) $A \cap B \in \cF$ &
    (iv) $A \setminus B \in \cF$ &
    (v) $B \setminus A \in \cF$
%   \end{enumerate}      
    \end{tabular}

    In other words, $\cF$ contains all the trivial
    subsets of $U$, $A \cup B$ and the partitions of $A \cup B$
    defined by intersection and set difference.

@@

@@

The following theorem gives a very interesting property of orthogonal
sets and $\overline{\cF}$

@@

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree, as alluded to by
Corollary~\ref{cor:pqrrecursive}. They are calculated as follows.
First method is by computing the overlap components of $\cF$. The
overlap components is the partition that results from the {\em
  overlap} equivalence relation which is nothing but {\em
  non-orthogonal} equivalence relation (It is easy to verify
  that this relation is indeed an equivalence relation.). Overlap
components are linearly computable\cite{mm95,wlh92}. Once these
elements are factored out, the rest of the node sets are obtained by
the second method which is identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {\em indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01} and even in
logspace\cite{kklv10}.

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96}}] %,\cite[Th.~2.1]{mcc04}
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$ for
  some permutation $\alpha$ on $U$ ($consect(\alpha)$ is a linear order)
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

\subsubsection{Comparision with generalized \PQtree}

Stepping back a little, the reader may have noticed the cross
references between \PQRtree and other data structures in footnotes
earlier.  Generalized \PQtree or \gPQtree is a data structure defined
in \cite{n89} to represent all orthogonal sets of a set system
$\cF$. A data structure with the same name was later defined in
\cite{mcc04} as part of a {\em substitution decomposition} for a set
system $\cF$ and subsequently \cite{mcc04} gives a new
characterization of $\cF$ using a so-called incompatibility graph
(this is discussed in more detail in
Section~\ref{sec:incompatibilitygr}). As seen above, \PQRtree is
defined in \cite{mm96} as a data structure to represent any set system
$\cF$ with additional information in their $R$-nodes if $\cF$ has no
\COP. All three of these data structures are proposed as
generalizations of \cite{bl76}'s \PQtree and hence produce the \PQtree
if $\cF$ has \COP. As a whole, all these three data strutures are
largely identical with their differences being notional. We discussed
the basic theory that they all hold and predominantly used the
terminology from \cite{mm96}. To the best of our knowledge this
comparison study has not been made earlier in the literature.  We
observe the three data strutures of \PQRtree, \gPQtree and generalized
\PQtree to be equivalent. The terminologies of the theory of all these
data structures is summarised in Table~\ref{tab:pqrcomparison}.

Going back to the theory we were discussing earlier,
Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[esp. Th.~2.1, 3.5. also Th.~3.2, 3.3, 3.4]{mcc04}
which categorizes the nodes in the above the three cases as {\em
  prime}, {\em linear} and {\em degenerate}
respectively. \cite{mcc04}'s generalized \PQtree is created in similar
ways as \PQRtree above. This tree is, in essence, the Hasse diagram of
what they call {\em strong elements} of weak closure, $\cW(\cF)$(\ie
$\overline{\cF}$). Strong elements of a set family are elements that
do not overlap with any other elements in the family, \ie it is
orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

@@

@@

    &
%     \parbox[t]{\colwidth}
%       {}
    \PQRtree
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
    }
    \\\hline

@@

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }
    &
    \parbox[t]{\colwidth}
    {
      {\em Prime}: For every $I$, $\emptyset \ne I \subset [k],
      \cup_{i \in I}S_i \notin \cF$\\
      {\em Linear}: There exists an ordering of $[k]$
      s.t. $\emptyset \ne I \subset [k]$ implies $\cup_{i \in I}S_i
      \in \cF$ only if members of $I$ are consecutive in the ordering\\
      {\em Degenerate}: For every $I$, $\emptyset \ne I \subset [k],
      \cup_{i \in I}S_i \in \cF$\\
    }

    \\\lasthline

@@

\subsection{ICPIA - a set cardinality based \COP test}
\label{sec:icpiasurvey}
In this section we see a characterization that does not use
\PQtree. \cite{nsnrs09} describes a characterization of \cop solely
based on the cardinality properties of the sets in the set system and
does not use any variants of \PQtrees.

@@

@@

\subsection{Finding forbidden submatrices}
\label{sec:forbiddensubalgo}

It is known from Theorem~\ref{th:tuckercop} that when a matrix does
not have \COP, it must have one of Tucker's forbidden submatrices (See
Section~\ref{sec:tucker} for detailed discussion on Tucker's forbidden
substructures for matrices with \COP). However, none of the \COP test
algorithms discussed in Section~\ref{sec:surveycoptest} outputs the
forbidden submatrix when the test is negative. This perhaps gives a
hint about how hard it is to find these substructures in the input
matrix. Using a reduction to the longest path problem, \cite{v85}
proves that finding a Tucker's forbidden submatrix of at least $k$
rows is NP complete. An example of an optimization problem for a
matrix without \COP is finding if deleting at most $k$ rows from input
will restore \COP. This is known to be NP complete
\cite{b75-phd}. There are several such optimization problems that try
to find ways to get \COP on the input matrix with certain alterations
and finding smallest forbidden submatrix is at the heart of these problems.

In this section we will summarize \cite{d08phd}'s approximation
algorithm for finding forbidden submatrix. We will use $\Tuckerset$ to
refer to the set of Tucker's forbidden submatrices shown in
Figure~\ref{fig:forbiddensubmatrices}.

@@

Using Corollary~\ref{cor:tuckertheorem}, a forbidden submatrix can be
found as follows. For every vertex triple $x, y, z$ in $G_M$ that
correspond to columns of $M$, determine the sum of the lengths of
three shortest paths connecting each pair of these vertices (there are
3 possible pairs) avoiding the neighborhood of third vertex. If all
three paths exist, then $x, y, z$ indeed form an asteroidal
triple. Among all such triples, select the triple $u, v, w$ where the
sum is minimum and return the rows and columns of $M$ that correspond
to the vertices in the three shortest paths described earlier for this triple. This
return value forms a submatrix and it obviously must contain a
submatrix from $\Tuckerset$ because it represents an asteroidal
triple. It may be noted that this procedure does not always return a
submatrix of smallest size because the shortest paths computed may not
be vertex disjoint. The following claim shows that this certainly
gives a reasonable approximation of the solution. The claim has been
abridged to summarize the main idea behind this algorithm.
Note that a $(x,y)$-matrix is one with at most $x$ number of {\un}s in every
column and at most $y$ number of {\un}s in every row. If $x$ or $y$ is given as
$*$, it denotes that number of {\un}s is unbounded.

\begin{proposition}[{\cite[Pr.~3.1, {\em abridged}]{d08phd}}] 
  \label{prop:tuckerdomapprox}
  Let $M$ be a $(*,\Delta)$-matrix of size $m \times n$ that contains
  a forbidden submatrix $M'$ of size $m' \times n'$ from $\Tuckerset$. Then
  this algorithm computes a submatrix that belongs to $\Tuckerset$
  with at most $m'+5$ rows and $n'+3$ columns in $O(\Delta m n^2 +
  n^3)$ time.
\end{proposition}

In fact, the approximation mentioned in
Proposition~\ref{prop:tuckerdomapprox} is the worst case when
the actual solution $M'$ is of type $M_{IV}$. The approximation is
slightly better for the other types of submatrices from $\Tuckerset$,
the best case being an exact solution when $M'$ is $M_{I_k}$ or
$M_{II_k}$. Thus this algorithm works accurately when the solution is of
type $M_{I_k}$ or $M_{II_k}$.

Next, notice that Tucker matrix types $M_{IV}$ and $M_{V}$ are
constant size -- $4 \times 6$ and $4 \times 5$ respectively. Thus they
can be searched for in polynomial time even with brute force method as
Corollary~\ref{cor:mivmvpolytime} states.

\begin{corollary}[{\cite[Cor.~3.3, {\em abridged}]{d08phd}}] 
  \label{cor:mivmvpolytime}
  Let $M$ be a binary $(*, \Delta)$-matrix of size $m \times n$ and
  $M'$ be a binary matrix from $\Tuckerset$. A submatrix in $M$ that
  is isomorphic to $M'$ can be found in $O(\Delta^3m^2n^3)$ time or
  $O(\Delta^4m^2n)$ time when $M'$ is of type $M_{IV}$ or $M_V$
  respectively.
\end{corollary}

Their algorithm for finding $M_{III_k}$ type submatrix is slightly
involved and we only present here the analysis result in
Proposition~\ref{cor:miiitime}. The algorithm is derived from the
structural similarities between the first three types of Tucker
matrices and the fact that $M_{I_k}$ is a chordless cycle of length
$2k+4$.

\begin{proposition}[{\cite[Pr.~3.6]{d08phd}}] 
  \label{cor:miiitime}
  Let $M$ be a $(*,\Delta)$-matrix $M$ of size $m \times n$. Then a
  minimum size submatrix of type $M_{III_k}$ in $M$ can be found in
  $O(\Delta^3m^3n+\Delta^2m^2n^2)$ time.
\end{proposition}

All the above algorithmic results are combined to obtain the final
single algorithm for finding forbidden
submatrix. Theorem~\ref{th:forbiddenalg} gives this result.

\begin{theoremsansproof}[{\cite[Th.~3.1]{d08phd}}]
  \label{th:forbiddenalg}
  Let $M$ be a $(*,\Delta)$-matrix of size $m \times n$. A forbidden
  submatrix from $\Tuckerset$ in $M$ that has a minimum number of rows
  can be found in $O(\Delta^3m^2n\dot(m+n^2))$ time. Within the same
  time, one can also find a forbidden submatrix from $\Tuckerset$ in
  $M$ that has a minimum number of columns, a minimum number of rows
  and columns, or a minimum number of entries.
\end{theoremsansproof}

The procedure used to prove Theorem~\ref{th:forbiddenalg} is as
follows. 
\begin{enumerate}
\item Run the algorithm mentioned in
  Propostition~\ref{prop:tuckerdomapprox} to obtain a submatrix $A$
  which is an approximation of the smallest forbidden submatrix. (If
  the actual solution is $M_{I_k}$ or $M_{II_k}$ this will be an exact
  solution.)
\item Run the algorithm from Propostition~\ref{cor:miiitime} to find
  an induced $M_{III_k}$ and let this be $B$.
\item Run the algorithm from Corollary~\ref{cor:mivmvpolytime} twice
  -- once each for finding $M_{IV}$ and finding $M_{V}$. Let the
  results be $C$ and $D$ respectively.
\item Return the minimum size matrix among $A, B, C, D$ (minimum size
  being one of the following: number of rows, number of columns, sum
  of number of rows and columns, number of {\un} entries).
\end{enumerate}

In the next section we see yet another forbidden substructure for
\COP which is easier to compute.

\subsection{Incompatibility graph - a certificate for no \COP}
\label{sec:incompatibilitygr}

@@

\cite{mcc04} describes another characterization for matrices without
\COP which is efficiently computable. This is done by way of an {\em
  incompatibility graph} of a set system (since \COP in
  matrices and set systems are equivalent problems as seen in
  Section~\ref{sec:surveycertalgo} we will use the term set system in
  this section, in line with \cite{mcc04}'s convention). The
construction of this graph is based on the following observation.

@@

\begin{definition}[{\em Incompatibility graph \cite[Def.~6.1]{mcc04}}]
  \label{def:incompatibility}
  Let $\cF$ be a set system with universe $U$. Let $A_{\cF} = \{(a,b)
  \mid a, b \in U, a \ne b\}$. The {\em incompatibility graph}
  $G_I^\cF$ of $\cF$ is an undirected graph defined as follows.
%  \singlespacing
  \begin{enumerate}
  \item   The vertex set of $G_I^\cF$ is $ A_{\cF}$
  \item  The edge set of $G_I^\cF$ are pairs of the following forms
    \begin{itemize}
    \item $\{(a,b),(b,a)\}$ for all $a, b \in U$
    \item $\{(a,b),(b,c)\}$ if there exists $S \in \cF$ \stt $a, c
      \in S$ and $b \notin S$, for all $a, b, c \in U$
    \end{itemize}
  \end{enumerate}
  \dstop
\end{definition}

Thus an edge between two vertices $(a,b), (b,c)$ in $G_I^{\cF}$ means
that there exists no linear order on $U$ that can place $b$ after $a$
and place $c$ after $b$.  For any \COP order, there must not be any
incompatible pairs. Thus it must consist of an independent set $I$ of
the incompatibility graph. Moreover the independent set must have
exactly half the vertex set of $G_I^\cF$ since the \COP order must
involve all elements in $U$. Secondly, the reverse of a \COP order is
also a \COP order. Thus $A_\cF \setminus I$ must also be an
independent set. In other words, $G_I^\cF$ must be bipartite with
partitions being of equal size if $\cF$ has \COP. An odd cycle is a
forbidden structure in a bipartite graph, hence the same is forbidden
in the incompatibility graph of a set system with \COP. In other
words, if the set system has no \COP, its incompatibility graph must
have an odd cycle.  There is also a requirement on the length of this
odd cycle which is formally stated in Theorem~\ref{th:mccbipartite}.
